Biology background

### 2014-11-08 Rmarkdown/biology_background.Rmd
Finished first draft of biology background chapter.

In this chapter I summarise the basic biology necessary for
understanding my thesis. I then introduce DNA methylation, describe some of its biological roles and discuss the types of questions asked in experiments studying DNA methylation. Finally, I describe assays for studying DNA methylation, with a particular focus on the whole-genome bisulfite-sequencing assay known as _methylC-seq_.

@@

A major reason that DNA methylation is studied is that it is essential for normal development in mammals \cite{Li:1992bp}. It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression (__CITE__), silencing of transposable elements (__CITE__), X chromosome inactivation (__CITE__) and tumorigenesis (__CITE__). However, DNA methylation is not found in all organisms, for example, it not detectable in yeast, and so is not essential for life. Furthermore, the level of DNA methylation varies widely amongst different organisms, with many having very low levels of methylation (__TODO: examples and citations__).

When people speak of DNA methylation they are generally referring to methylation of the cytosine nucleobase. Even more specifically, they are referring to the molecule _5-methylcytosine_. 5-methylcytosine, abbreviated as _5mC_, is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm]. However, I will continue to use the terms _DNA methylation_ and _cytosine methylation_ interchangeably, as is standard in the literature. The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \cite{@Lister:2009boa}.

@@

A German chemist, W.G. Ruppel, first identified a methylated nucleic acid in 1898. Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide (__TODO: Can only find first page of this paper and it's in German. Need to verify__). In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \cite{Johnson:1925js}. However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \cite{@VISCHER:1949ty}.

@@

DNA methylation is essential for normal development in mammals, such as human and mouse (__REFERENCE__). My thesis focuses on DNA methylation in humans but much of the material in this section is true for other mammals.

@@

In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_. A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence (__FIGURE__). The "p" stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation. A CpG is a palindromic sequence and is generally symmetrically methylated. Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the dinculeotide is a 5-methylcytosine (__REFERENCE__).

@@

One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \cite{@Scarano:1967ue}. Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see __FIGURE__).

@@

This definition was refined by \cite{Takai:2002jb} to exclude _Alu_-repetitive elements, which otherwise are misclassified as _bona fide_ CpG islands. More recently, \cite{@Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.

@@

- CpG island shores: regions within 2 kb of CpG islands \cite{Irizarry:2008hg}. CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}.
- CpG island shelves have been defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.
- The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \cite{Sandoval:2011dj}.

@@

Overall, non-CpG methylation is less well studied and less well understood than CpG methylation. This is partly due to sampling bias - commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k beadchips (__TODO: ref section on technology__), only measure CpG methylation.  However, recent technological advances (__TODO: ref section on technology__) mean that cytosine methylation can be routinely assayed regardless of the sequence context.

@@

\subsection{Function of DNA methylation: "normal", cancer and development}\label{sec:function_of_dna_methylation}

- [ ] DONE

\subsection{DNA methylation in other organisms}\label{sec:dna_methylation_in_other_organisms}

- [ ] DONE

__TODO: Don't want too much detail here__

\subsubsection{Insect}

- [ ] DONE

* Honeybees
* Fruit fly? DNA methylation's existance in other organisms, such as the common fruit fly (_Drosophila Melanogaster_), is an area of contention although recent studies suggest it is present albeit at very small quantities and only in very limited contexts [@Takayama:2014jp].
* General

\subsubsection{Plants}

- [ ] DONE

* Arabadopsis
* Non-CG methylation

\subsubsection{Fungi}

- [ ] DONE

\subsubsection{Bacteria}

- [ ] DONE

\subsubsection{Archaea}

- [ ] DONE

@@

In the 1980s, two research groups independently discovered that when DNA is treated with sodium bisulfite ($NaHSO_{3}$), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines \cite{Shapiro:1970im, Hayatsu:1970jv}. The methylated cytosines are said to be "protected" from conversion to uracils. This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite \cite{Frommer:1992wq}, which are referred to as bisulfite-conversion assays.

@@

\subsubsection{Analysis platforms combined with bisulfite conversion}\label{sec:bs_platforms}

__UP TO HERE__

- [ ] DONE

@@

Microarrays contain thousands, even millions, of short _oligonucleotide_ probes. Each probe is is designed to hybridise to a particular DNA sequence and emits a signal, such as flurousence, that can be measured to infer the strength of the hybridisation. Therefore, an (idealised) way to analyse DNA methylation with a microarray is to hybridise bisulfite-converted DNA to a microarray that contains probes for both the methylated and unmethylated versions of all sequences of interest. The relative methylation of each sequence can be inferred from the relative intensities of the "methylated probe" to the "unmethylated probe". Such an idealised experiment brushes over many complications including [@Laird:2010iv]:

* The reduced complexity of bisulfite-converted DNA (from a 4-base alphabet to a largely 3-base alphabet) leads to decreased hybridisation specificity
* Sequences containing multiple cytosines require multiple probe versions in order to assay all possible methylation patterns
* This approach requires the design of organism-specific microarrays

The Illumina 450k (__INSERT FULL NAME__) provides a modern implementation of the idea of bisulfite pre-treatment of DNA followed by microarray hybridisation for studying human samples. This array assays __HOW MANY__ CpGs (__IN WHAT TYPE OF REGIONS__) and uses two different probe designs (__WHY?__). There are still substantial bioinformatics challenges in analysing data from these assays (__CITE: SNPs in probes, normalisation, etc.__).

As mentioned in __SECTION__, genomics research was revolutionised by the development of cheap high-throughput sequencing technology and the study of DNA methylation was no exception. In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Illumina sequencing technology [@Cokus:2008fc, Lister:2008bh]. [@Cokus:2008fc] termed their approach _BS-seq_ while [@Lister:2008bh] called theirs _methylC-seq_. These techniques, along with methods that these with targeting of particular regions of the genome, are described further in __SECTION__.

One final approach to mention is the use of mass spectrometry to analyse bisulfite-treated DNA, as implemented in Sequenom's EpiTYPER (__SOURCE__). This platform can provide quantitative measurements of CpG methylation across hundreds of loci and multiple samples and is frequently used to validate findings discovered using other platforms (__SOURCE__).

- [ ] DONE

\subsection{Pros and cons of bisulfite-conversion assays}\label{sec:pros_and_cons_of_bs_assays
}

- [ ] DONE

Bisulfite-conversion assays are considered the "gold standard" for studying DNA methylation. Currently, multiple cells are required as input although the minimum number required is being lowered and efforts are well produce single-cell, whole-genome DNA methylation maps using bisulfite-treated DNA (__SOURCE__).

@@

A recently discovered disadvantage is that bisulfite-conversion assays are unable to distinguish between 5mC and 5hmC because both modifications similarly protect cytosine from deamination to uracil (__SEE [@Yu:2012ej] for references__). In effect, the detection of 5mC, and all subsequent inference, is confounded with that of 5hmC. For most experiments this isn't much of a problem - most cells have very low levels of 5hmC and so there is little confounding - however, in certain experiments this needs a more careful approach. To address this issue, [@Booth:2012fh] developed _oxidative bisulfite-sequencing_ (oxBS-seq) and [@Yu:2012ej] developed _Tet-assisted bisulfite sequencing_ (TAB-seq) to separate 5mC and 5hmC detection.

oxBS-seq specifically measures 5mC. The input DNA is oxidated by potassium perruthenate (KRuO$_{4}$), which converts 5hmC to 5fC, prior to bisulfite-treatment. Only 5mC is protected from conversion during the bisulfite-treatment, which effectively means that only 5mC remains to be detected at the sequencing stage. The level of 5hmC can be estimated by performing traditional bisulfite-sequencing and then "subtracting" the oxBS-seq signal (5mC) from the bisulfite-sequencing signal (5mC + 5hmC).

@@

A disadvantage of bisulfite-conversion assays is that they require knowledge of the underyling DNA sequence in order to infer the methylation states of cytosines. This requires either a parallel experiment to sequence the target region(s) or reliance on a reference genome. When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample (__SOURCE__). __FIGURE__ describes such an example.

The bisulfite-treatment of DNA can introduce biases and other problems. Three examples are _PCR-bias_, _incomplete bisulfite conversion_ and _DNA degradation_. _PCR-bias_ is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence [@Warnecke:1997eh]. _Incomplete bisulfite-conversion_ leads to cytosines being incorrectly inferred as 5mC (__SOURCE__). Finally, bisulfite-treatment can lead to _DNA degredation_, which, as the name suggets, leads to a loss in DNA quality or quantity and can make it particularly hard to sequence long DNA molecules (__SOURCE: Anecdotal__).

\section{Whole-genome bisulfite-sequencing}\label{sec:wgbs}

- [ ] DONE

\subsection{Targeted BS-seq, e.g. RRBS, NimbleGen and Agilent's capture kits, MRE-seq}\label{sec:targeted_bsseq}

- [ ] DONE

* "RRBS’s ability to interrogate a locus is dependent on its MspI-cut-site (CCGG) density and consequently measures 10%–15% of the CpGs in the human genome (Bock et al. 2010; Harris et al. 2010)" \cite{Stevens:2013hv}
* See \cite{Stevens:2013hv} for MRE-seq details and a good summary of DNA methylation assays

>>>>

In this chapter I summarise the basic biology necessary for
understanding my thesis. I introduce DNA methylation, describe some of its biological roles and assays for studying DNA methylation, with a particular focus on bisulfite-sequencing.

@@

A major reason that DNA methylation is studied is that it is essential for normal development in mammals \cite{Li:1992bp}. It is also involved in many key biological processes related to human development and health, such as the regulation of gene expression (__CITE__), silencing of transposable elements (__CITE__), X chromosome inactivation (__CITE__) and tumorigenesis (__CITE__).

However, DNA methylation is not found in all organisms, for example, it not detectable in yeast, and so is not essential for life. Furthermore, the level of DNA methylation varies widely amongst different organisms, with many having very low levels of methylation (__TODO: examples and citations__).

When people speak of DNA methylation they are generally referring to methylation of the cytosine nucleobase. Even more specifically, they are referring to the molecule _5-methylcytosine_. 5-methylcytosine, abbreviated as _5mC_, is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm]. However, I will continue to use the term _DNA methylation_ to describe the more specific 5mC, as is standard in the literature.

@@

A German chemist, W.G. Ruppel, first identified a methylated nucleic acid in 1898. Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide (__TODO: Can only find first page of this paper and it's in German. Need to verify__). In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \cite{Johnson:1925js}. However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \cite{VISCHER:1949ty}.

@@

@@

- [x] DONE
The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \cite{Lister:2009boa}. In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_. A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence (__FIGURE__). The "p" stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation. A CpG is a palindromic sequence and is generally symmetrically methylated. Approximately 60-80% of CpGs are methylated in mammals, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine (__REFERENCE__).

@@

One reason for this relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \cite{Scarano:1967ue}. Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see __FIGURE__).

@@

This definition was refined by \cite{Takai:2002jb} to exclude _Alu_-repetitive elements, which otherwise are misclassified as _bona fide_ CpG islands. More recently, \cite{Wu:2010do} developed a hidden Markov model to predict CpG islands based on the CpG density and GC-content of the region.

@@

- CpG island shores: regions within 2 kb of CpG islands \cite{Irizarry:2008hg}. CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}. Shelves have a comparatively low density of CpGs.
- CpG island shelves have been defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.
- The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \cite{Sandoval:2011dj}.

@@

Overall, non-CpG methylation in humans is less well studied and less well understood than CpG methylation. This is partly due to sampling bias - commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k beadchips (__TODO: ref section on technology__), only measure CpG methylation.  However, recent technological advances (__TODO: ref section on technology__) mean that cytosine methylation can be routinely assayed regardless of the sequence context.

Non-CpG methylation is very common in other organisms, such as plants. For example, \cite{Lister:2008bh} found that $45\%$ of "methylcytosines" (see __SECTION__ for the definition and a criticism) in the widely-studied _Arabidopsis Thaliana_ are at CHG or CHH loci. However, they also found that the level of methylation at non-CpG loci is typically lower than that observed at CpG dinucleotides.

@@

@@

In the 1980s, two research groups independently discovered that when DNA is treated with sodium bisulfite ($NaHSO_{3}$), unmethylated cytosines deaminate to uracils much faster than do methylated cytosines \cite{Shapiro:1970im, Hayatsu:1970jv}. The methylated cytosines are said to be "protected" from conversion to uracils. This discovery led to the development of assays for studying studying cytosine methylation based on the pre-treatment of DNA with sodium bisulfite \cite{Frommer:1992wq, Clark:1994wc}, which are referred to as bisulfite-conversion assays.

@@

@@

Microarrays contain thousands, even millions, of short _oligonucleotide_ probes. Each probe is is designed to hybridise to a particular DNA sequence and emits a fluorescent signal that can be measured to infer the strength of the hybridisation. Therefore, an (idealised) way to analyse DNA methylation with a microarray is to hybridise bisulfite-converted DNA to a microarray that contains probes for both the methylated and unmethylated versions of all sequences of interest. The relative methylation of each sequence can be inferred from the relative intensities of the "methylated probe" to the "unmethylated probe". Such an idealised experiment brushes over many complications including \cite{Laird:2010iv}:

- The reduced complexity of bisulfite-converted DNA (from a 4-base alphabet to a largely 3-base alphabet) leads to decreased hybridisation specificity
- Sequences containing multiple cytosines require multiple probe versions in order to assay all possible methylation patterns
- This approach requires the design of organism-specific microarrays

The Illumina Infinium HumanMethylation450 BeadChip (_Illumina 450k_) provides a modern implementation of this type of assay for studying human DNA. This array assays $482,422$ cytosines, $99.3\%$ of these CpGs, across a wide variety of genomic features \cite{Stirzaker:2014cu}. There are well-established methods and software for analysing Illumina 450k data such as `minfi` \cite{Aryee:2014by}

Genomics research was revolutionised by the development of cheap high-throughput sequencing technology, and the study of DNA methylation was no exception. In 2008, two papers were published describing methods for whole-genome shotgun sequencing of bisulfite-converted DNA using the nascent Illumina sequencing technology \cite{Cokus:2008fc, Lister:2008bh}. Whole-genome bisulfite-sequencing is the gold-standard for analysing genome-wide DNA methylation data.

\cite{Cokus:2008fc} termed their approach _BS-seq_ while \cite{Lister:2008bh} called theirs _methylC-seq_. From a bioinformatics perspective, the main difference is that the _BS-seq_ protocol produces sequencing reads from four bisulfite-converted DNA strands - the original top strand (_OT_), the complementary strand to the original top strand (_CTOT_), the original bottom strand (_OB_) and the complementary strand to the original bottom strand (_CTOB_) - which require mapping to four different _in silico_ converted reference genomes, followed by a merge of the alignment results. In contrast, the _methylC-seq_ protocol only produces sequencing reads from two bisulfite-converted strands - _OT_ and _OB_ - and so only requires mapping to two different _in silico_ converted reference genomes, followed by a merge. The methylC-seq protocol is now the standard whole-genome bisulfite-sequencing protocol, due in part to the simpler bioinformatics analysis.

Whole-genome bisulfite-sequencing remains an expensive assay, which limits its use in studies involving large numbers of samples. Furthermore, approximately $40\%$ of reads will not contain any CpGs, which, depending on the experiment, might be considered a gross waste of money. Several assays have been developed to perform targeted high-throughput bisulfite-sequencing. The targeted nature of these assays are their obvious advantage and disadvantage - only a subset of the genome needs be sequenced but you only obtain information about methylation in those regions. Depending on the experiment this tradeoff may be worthwhile, and these assays have been successfully used in a number of studies (__CITE__).

Some examples of these targeted bisulfite-sequencing are:
- Reduced representation bisulfite-sequencing (_RRBS_, \cite{Meissner:2005eb}): RRBS uses restriction enzymes to first select regions of the genome with a high CpG density (based on the Msp-I cleavage motif), which are subsequently treated with sodium-bisulfite and sequenced.
- NimbleGen's SeqCap Epi Enrichment System and Agilent's SureSelectxt Human Methyl-Seq: Both of these commercial products use DNA hybridisation to enrich the sequencing library for pre-defined regions of interest. This enriched library is then bisulfite-converted and sequenced.

One final class of bisulfite-conversion assays does not use microarrays or high-throughput sequencing. Sequenon's EpiTYPER \cite{Ehrich:2006tt} uses mass spectrometry to analyse DNA methylation from bisulfite-converted DNA. This platform can provide quantitative measurements of CpG methylation across hundreds of loci and multiple samples and may be used to validate findings discovered using other platforms \cite{Laird:2010iv}.

\subsubsection{Pros and cons of bisulfite-conversion assays}\label{sec:pros_and_cons_of_bs_assays
}

Bisulfite-conversion assays are considered the "gold standard" for studying DNA methylation since cytosine methylation can be detected at single-base resolution \cite{Stirzaker:2014cu}. However, almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1-5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells. The minimal amount of DNA is being reduced with each technological advance. Furthermore, single-cell bisulfite-sequencing assays have recently been reported \cite{Guo:2013ih, Smallwood:2014kn}[^not_genome_wide]

[^not_genome_wide]: These are not yet proper genome-wide assays. The technique of \cite{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome. And the technique of \cite{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to 48.4% of cpG sites" \cite{Smallwood:2014kn}.

@@

A recently discovered disadvantage is that bisulfite-conversion assays are unable to distinguish 5hmC from 5mC because hydroxymethyl similarly protects a cytosine from deamination to uracil. In effect, the detection of 5mC, and all subsequent inference, is confounded with that of 5hmC. For most experiments this isn't much of a problem - most cells have very low levels of 5hmC and so there is little confounding - however, in certain experiments this needs a more careful approach. To address this issue, \cite{Booth:2012fh} developed _oxidative bisulfite-sequencing_ (oxBS-seq) and \cite{Yu:2012ej} developed _Tet-assisted bisulfite sequencing_ (TAB-seq) for separate 5mC and 5hmC detection.

oxBS-seq specifically measures 5mC. The input DNA is oxidated by potassium perruthenate (RKRuO_{4}$), which converts 5hmC to 5fC, prior to bisulfite-treatment. Only 5mC is protected from conversion during the bisulfite-treatment, which effectively means that only 5mC remains to be detected at the sequencing stage. The level of 5hmC can be estimated by performing traditional bisulfite-sequencing and then "subtracting" the oxBS-seq signal (5mC) from the bisulfite-sequencing signal (5mC + 5hmC).

@@

There is also potential confounding of 5C, 5mC and 5hmC levels with 5fC and 5caC, although this has received less attention since 5fC and 5caC are believed to exist at far lower quantities than 5mC and 5hmC. Nonetheless, sequencing assays of 5fC and 5caC exist based on the idea of treating the DNA with a chemical and performing 'signal subtraction' with traditional bisulfite-sequencing or another assay \cite{Wu:2014jj}.

Another disadvantage of bisulfite-conversion assays is that they require knowledge of the underlying DNA sequence in order to infer the methylation states of cytosines. This requires either a parallel experiment to sequence the target region(s) or reliance on a reference genome. When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample \cite{Liu:2012ge}. __FIGURE__ illustrates such an example.

The bisulfite-treatment of DNA can introduce biases and other problems. Three examples are _PCR-bias_, _incomplete bisulfite conversion_ and _DNA degradation_. _PCR-bias_ is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \cite{Warnecke:1997eh}. _Incomplete bisulfite-conversion_ leads to cytosines being incorrectly inferred as 5mC. Finally, treating DNA with sodium bisulfite damages the DNA, resulting in the fragmentation of long molecules. This limits the size of the fragments that can be studied using bisulfite-conversion assays to $\approx 500$ bp \cite{Ecker:2010et}

Ideally, an assay of DNA methylation could naturally 'read' methylated cytosines as a distinct signal from unmethylated cytosines. This idea forms the basis of assays using the kinetics of DNA to infer the presence of DNA methylation and other DNA modifications. While very exciting, these assays do not yet scale to studying genome-wide DNA methylation levels in mammalian-sized genomes.

The only commercially available assay in this class is the Pacific Biosciences SMRT technology \cite{Flusberg:2010hra}. It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a nucleotide, such as 5mC, to the time it takes to read its unmodified form.
